En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Niccol\`o Veltri was supported by a research grant (13156) from VILLUM FONDEN.}

### 2019-04-30 mpc19/paper.tex
Homogenized our use of \Delay vs. delay, \Maybe vs. maybe, \trace vs. trace.

In this paper, we extend the two-level calculus of finitary \Pio{} to one for
full \Pio{} (i.e., with both recursive types and iteration by means of a
\trace{} combinator) using the \Delay{} monad, which can be regarded as a
``computability-aware'' analogue of the usual $\mathsf{Maybe}$ monad for
partiality. This yields a calculus of iterative (and possibly non-terminating)
reversible programs acting on user-defined dynamic data structures together
with a calculus of certified program equivalences between these programs.

@@

In \cite{CaretteS16}, the authors studied the equivalences of isomorphisms in
the finitary fragment of \Pio{} (i.e., without recursive types and iteration
via the \trace{} combinator), and showed that these equivalences could be
adequately described by another combinator calculus of equivalences of
isomorphisms, in sum yielding a two-level calculus of isomorphisms and
equivalences of isomorphisms. In this paper, we build on this work to produce a
(fully formalized) two-level calculus for full \Pio{} (supporting both recursive
types and iteration) via the \Delay{} monad, using insights gained from the
study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well
as of join inverse categories in which reversible iteration may be
modelled~\cite{KaarsgaardAG17}.

@@

The delay datatype is a monad. The unit is the constructor $\now$,
while the Kleisli extension $\bind$ is corecursively defined as follows:
\begin{align*}
& \bind : (A \to \Delay B) \to \Delay A \to \Delay B \\
& \bind \,f \, (\now\,a) = f\,a \\
& \bind\,f\,(\later\,x) = \later\,(\bind\,f\,x)
\end{align*}
The delay monad, like any other monad on $\Set$, has a unique strength
operation which we denote by $\str : A \times \Delay B \to \Delay (A
\times B)$. Similarly, it has a unique costrength operation $\costr :
\Delay A \times B \to \Delay (A \times B)$ definable using $\str$. Moreover, the delay datatype is a commutative monad.

@@

\section{Conclusions}
In this paper, we have extended the work of Carette and Sabry~\cite{CaretteS16}
to a (fully formalized) two-level calculus of \Pio{} programs and program
equivalences. Key in this effort was the use of the Kleisli category of the
$\Delay$ monad on $\Set$ under weak bisimilarity, which turned out to support
iteration via a trace that preserves all partial isomorphisms, in this way
giving semantics to the dagger trace of \Pio. Further, the work was formalized
using Agda 2.6.0.

It is natural to wonder if our work can be ported to other monads of partiality
in Martin-L\"of type theory. As already discussed in the end of Section
\ref{sec:trace}, the $\Maybe$ monad is not suitable for modelling a
well-behaved trace combinator without the assumption of classical principles
such as \textsf{LPO}. The partial map classifier \cite{UustaluV17-2,EscardoK17}
$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where $\mathsf{Prop}$ is
the type of propositions (types with at most one inhabitant), supports the
existence of a uniform iteration operator and therefore a trace. Nevertheless,
the specification of iteration is more complicated then the one presented in
Section \ref{sec:elgot} for the delay monad, which is a simple corecursive
definition. The Elgot monad structure of $\mathsf{PMC}$ follows from its
Kleisli category being a join restriction category, so iteration is defined in
terms of least upper bounds of certain chains of morphisms. The subcategory of
partial isomorphisms of the Kleisli category of $\mathsf{PMC}$ supports a
dagger trace combinator, which can be proved following the general strategy in
\cite{KaarsgaardAG17}. The exact same observations apply to the partiality
monad in homotopy type theory \cite{AltenkirchDK17,ChapmanUV19}, to which the
quotiented delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
isomorphic to under the assumption of countable choice.

Though the Kleisli category of the $\Delay$ monad on $\Set$ is well studied,
comparatively less is known about this monad on other categories. It could be
interesting to study under which conditions its iterator exist -- e.g., whether
this is still the case when $\Set$ is replaced with an arbitrary topos. Another
avenue concerns the study of \emph{time invertible} programming languages:
Though not immediately clear in the current presentation, the trace on
$\Inv\Dapprox$ is not just reversible (in the sense that it preserves partial
isomorphisms) but in fact time invertible, in the sense that the number of
computation steps needed to perform $\trace (\dagger f)$ is \emph{precisely}
the same as what is needed to perform $\trace~f$ on any input. Since the
$\Delay$ monad conveniently allows the counting of computation steps, we
conjecture that this is an ideal setting in which to study such intentional
semantics of reversible programming languages.

>>>>

In this paper, we extend the two-level calculus of finitary \Pio{} to one for
full \Pio{} (i.e., with both recursive types and iteration by means of a
\trace{} combinator) using the delay monad, which can be regarded as a
``computability-aware'' analogue of the usual maybe monad for
partiality. This yields a calculus of iterative (and possibly non-terminating)
reversible programs acting on user-defined dynamic data structures together
with a calculus of certified program equivalences between these programs.

@@

In \cite{CaretteS16}, the authors studied the equivalences of isomorphisms in
the finitary fragment of \Pio{} (i.e., without recursive types and iteration
via the trace combinator), and showed that these equivalences could be
adequately described by another combinator calculus of equivalences of
isomorphisms, in sum yielding a two-level calculus of isomorphisms and
equivalences of isomorphisms. In this paper, we build on this work to produce a
(fully formalized) two-level calculus for full \Pio{} (supporting both recursive
types and iteration) via the delay monad, using insights gained from the
study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well
as of join inverse categories in which reversible iteration may be
modelled~\cite{KaarsgaardAG17}.

@@

The delay datatype is a monad. The unit is the constructor $\now$,
while the Kleisli extension $\bind$ is corecursively defined as follows:
\begin{align*}
& \bind : (A \to \Delay B) \to \Delay A \to \Delay B \\
& \bind \,f \, (\now\,a) = f\,a \\
& \bind\,f\,(\later\,x) = \later\,(\bind\,f\,x)
\end{align*}
The delay monad, like any other monad on $\Set$, has a unique strength
operation which we denote by $\str : A \times \Delay B \to \Delay (A
\times B)$. Similarly, it has a unique costrength operation $\costr :
(\Delay A) \times B \to \Delay (A \times B)$ definable using $\str$. Moreover, the delay datatype is a commutative monad.

@@

\section{Conclusions}
In this paper, we have extended the work of Carette and Sabry~\cite{CaretteS16}
to a (fully formalized) two-level calculus of \Pio{} programs and program
equivalences. Key in this effort was the use of the Kleisli category of the
delay monad on $\Set$ under weak bisimilarity, which turned out to support
iteration via a trace that preserves all partial isomorphisms, in this way
giving semantics to the dagger trace of \Pio. Further, the work was formalized
using Agda 2.6.0.

It is natural to wonder if our work can be ported to other monads of partiality
in Martin-L\"of type theory. As already discussed in the end of Section
\ref{sec:trace}, the maybe monad is not suitable for modelling a
well-behaved trace combinator without the assumption of classical principles
such as \textsf{LPO}. The partial map classifier \cite{UustaluV17-2,EscardoK17}
$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where $\mathsf{Prop}$ is
the type of propositions (types with at most one inhabitant), supports the
existence of a uniform iteration operator and therefore a trace. Nevertheless,
the specification of iteration is more complicated then the one presented in
Section \ref{sec:elgot} for the delay monad, which is a simple corecursive
definition. The Elgot monad structure of $\mathsf{PMC}$ follows from its
Kleisli category being a join restriction category, so iteration is defined in
terms of least upper bounds of certain chains of morphisms. The subcategory of
partial isomorphisms of the Kleisli category of $\mathsf{PMC}$ supports a
dagger trace combinator, which can be proved following the general strategy in
\cite{KaarsgaardAG17}. The exact same observations apply to the partiality
monad in homotopy type theory \cite{AltenkirchDK17,ChapmanUV19}, to which the
quotiented delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
isomorphic to under the assumption of countable choice.

Though the Kleisli category of the delay monad on $\Set$ is well studied,
comparatively less is known about this monad on other categories. It could be
interesting to study under which conditions its iterator exist -- e.g., whether
this is still the case when $\Set$ is replaced with an arbitrary topos. Another
avenue concerns the study of \emph{time invertible} programming languages:
Though not immediately clear in the current presentation, the trace on
$\Inv\Dapprox$ is not just reversible (in the sense that it preserves partial
isomorphisms) but in fact time invertible, in the sense that the number of
computation steps needed to perform $\trace (\dagger f)$ is \emph{precisely}
the same as what is needed to perform $\trace~f$ on any input. Since the
delay monad conveniently allows the counting of computation steps, we
conjecture that this is an ideal setting in which to study such intentional
semantics of reversible programming languages.
